Forcing for Isabelle

### 2018-02-13 blueprint.tex
first part of the base case of forcing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}
The proofs are mostly based on Chapter IV of
Kunen's~\cite{kunen2011set}.

\section{Forcing notions}
We will need the following concepts:
\begin{itemize}
\item Forcing poset or \emph{notion}: a triple $\lb \PP, \1, \leq\rb$ where $\leq$ is a
  partial (pre)order on $\PP$ with (a distinguished) top element $\1$.
\item Compatible elements. Antichain in $\PP$. 
\item Dense subset $D\sbq\PP$ of a forcing poset.
\item Filter $G\sbq\PP$ in a poset.
\item $\calD$-generic filter for $\PP$, where $\calD$ is a family of
  dense subsets of $\PP$.
\end{itemize}

@@

As an example, consider $\PP,R \in M$,  and $R\sbq
\PP\times\PP$. Reflexivity of the relation $R$ is 
stated as follows.
\[
\phi\defi \forall p (p\in \PP \limp \lb p,p\rb \in R).
\]
In Isabelle/ZF, we write
%
\begin{verbatim}
definition 
  reflexivity_abs :: "[i,i] ⇒ o" where
  "reflexivity_abs(P,r) == ∀p . p∈P ⟶ <p,p>∈r"
\end{verbatim}

This is different from what would $M$ see from ``within.'' In
particular, the universal quantifier should range only over elements
of $M$:
\[
\phi^M\defi\forall p\in M (p\in \PP \limp \lb p,p\rb \in R).
\]
In Isabelle/ZF, we will move to a more general setting, by allowing
 $M$ to be a class.
\begin{verbatim}
definition  
  reflexivity_rel :: "[i⇒o,i,i] ⇒ o" where
  "reflexivity_rel(M,P,r) == ∀p[M].  p∈P ⟶ <p,p>∈r"
\end{verbatim}
(Notice: There must be no spaces between \verb|[M]| and the following
period.) This is the relativized version of
\verb|reflexivity_abs|.

@@

The last lemma shows that indeed we have defined an element of the set
\formula. The formula \verb|reflexivity_fm| is an internalization of
the Isabelle/ZF formula \verb|reflexivity_abs|, not the relativized
one.

@@

We will now make more precise the definition of the map
$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
$\phi'$ will have $n+4$ free variables, where the first four account
for the forcing notion and a particular condition. If
$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
\[
p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n).
\]
Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. 
%% For the
%% time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
%% variables. 
Informally, we call $\forces$ the \emph{forcing relation}.
Given $\quine\phi\in\formula$, we will write the forcing relation in
Isabelle/ZF by using  
\[
\forceisa(\quine\phi,\PP,\leq,\1,p)
\]

The base case (for atomic formulas) is famously the most
complicated one. 
\begin{multline}
  \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q)\lsii 
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q),
\end{multline}
\begin{multline}
  \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  \forceisa(\quine{t_1 =s},\PP,\leq,\1,q))\} \\
  \text{is dense below $p$.''}
\end{multline}

Actually, these cases depend on the theorem of well-founded recursion,
so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
must be obtained from the proof of that theorem. This means that we need
an internalized version of the theorem.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}
The proofs are mostly based on Chapter IV of
Kunen's~\cite{kunen2011set}, and the implementation will make use of
the work by Paulson~\cite{paulson_2003}.

\section{Forcing notions}
We will need the following concepts:
\begin{itemize}
\item Forcing poset or \emph{notion}: a triple $\lb \PP, \1, \leq\rb$ where $\leq$ is a
  partial (pre)order on $\PP$ with (a distinguished) top element $\1$.
\item Compatible elements. Antichain in $\PP$. 
\item Dense subset $D\sbq\PP$ of a forcing poset.
\item Filter $G\sbq\PP$ in a poset.
\item $\calD$-generic filter for $\PP$, where $\calD$ is a family of
  dense subsets of $\PP$.
\end{itemize}
(Check \verb|ZF/Order.thy| before doing any work.)

@@

As an example, consider $\PP,R \in M$,  and $R\sbq
\PP\times\PP$. Reflexivity of the relation $R$ is 
stated as follows.
\[
\phi\defi \forall p (p\in \PP \limp \lb p,p\rb \in R).
\]
In Isabelle/ZF, we write
%
\begin{verbatim}
definition 
  reflexivity :: "[i,i] ⇒ o" where
  "reflexivity(P,r) == ∀p . p∈P ⟶ <p,p>∈r"
\end{verbatim}

This is different from what would $M$ see from ``within.'' In
particular, the universal quantifier should range only over elements
of $M$:
\[
\phi^M\defi\forall p\in M (p\in \PP \limp \lb p,p\rb \in R).
\]
In Isabelle/ZF, we will move to a more general setting, by allowing
 $M$ to be a class.
\begin{verbatim}
definition  
  reflexivity_rel :: "[i⇒o,i,i] ⇒ o" where
  "reflexivity_rel(M,P,r) == ∀p[M].  p∈P ⟶ <p,p>∈r"
\end{verbatim}
(Notice: There must be no spaces between \verb|[M]| and the following
period.) This is the relativized version of
\verb|reflexivity|.

@@

The last lemma shows that indeed we have defined an element of the set
\formula. The formula \verb|reflexivity_fm| is an internalization of
the Isabelle/ZF formula \verb|reflexivity|, not the relativized
one.

@@

We will now make more precise the definition of the map
$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
$\phi'$ will have $n+4$ free variables, where the first four account
for the forcing notion and a particular condition. If
$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
\[
p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n).
\]
Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. 
%% For the
%% time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
%% variables. 
Given $\quine\phi\in\formula$, we will write this  in
Isabelle/ZF by using  
\[
\forceisa(\quine\phi,\PP,\leq,\1,p)
\]
Afterwards, the \emph{forcing relation} $\forces$ is defined by
relativizing/interpreting $\forces^*$ in a ctm $M$, for fixed
$\lb\PP,\leq,\1\rb\in M$:
\[
p\forces \phi(t_1,\dots,t_n) \defi 
\bigl(p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n)\bigr)^M\!.
\]

\subsection{Forcing for atomic formulas}
The base case of the definition of the forcing relation, that is, for
atomic formulas, is (in)famously the most complicated one. 
The definition adopted by Kunen~\cite[p.~257]{kunen2011set} is the
following, with two mutually recursive clauses:
\begin{multline}
  \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q)\lsii 
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q),
\end{multline}
\begin{multline}
  \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  \forceisa(\quine{t_1 =s},\PP,\leq,\1,q))\} \\
  \text{is dense below $p$.''}
\end{multline}
Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
to have a single, more complicated, clause for $t_1=t_2$ using
well-founded recursion on $t_1$
and $t_2$ and then define the case $t_1\in t_2$ explicitly.

In any case, these definitions depend on the theorem of well-founded
recursion,
so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
must be obtained from the proof of that theorem. Below, a defined
relation $R$ is \emph{set-like} if $R^{-1}\{a\}$ is a set for every $a$.
%
\begin{theorem}[{\cite[I.9.11]{kunen2011set}}]
  Assume that $R$ is a
  set-like well-founded defined relation on a class $A$, assume that 
  $\forall  x,s\,\exists!y\,\phi(x,s,y)$ and write $G(x,s)$ for the
  only $y$ such that $\phi(x,s,y)$. Then there exists another formula 
   $\psi$ such that $\forall x\,\exists!y\,\psi(x,y)$ and if we write
  $F(x)$ for the only $y$ satisfying $\psi(x,y)$, we have
  \[
  \forall a: F(a) = G(a,F\restr R^{-1}\{a\}).
  \]
\end{theorem}
%
Observe that since $F$ is a definable function, its restriction to the
set $R^{-1}\{a\}$ must also be a set by the Replacement Axiom.

There are several steps to apply this theorem to justify the
definition of forcing for atomic formulas:
\begin{enumerate}
\item As mentioned before, $F$ will be the characteristic function of
  the predicate $\forceisa$.
  %
\item The class $A$ is $V\times V = \{\lb x,y\rb : x,y\in V\}$,
  i.e.\ the class of all ordered pairs. The relation $R$ must be
  defined.
  %
\item Write down explicitly $\psi$ as an element of
  $\formula$. 
\end{enumerate}
%
Luckily, the definition of $R$ and the formula $\psi$ are spelled out
in Kunen's.
